---
title: "Homework 4"
author: "Ashley Sackpraseuth"
date: "10/4/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Flight Data

## Loading libraries
```{r}
library(nycflights13)
library(tidyverse)
```


## Questions

1. Find if each flight number is used only once in a day. (Hint: You may need the any() function).

```{r}
#Make a separate dataframe containing only the flight, year, month, and day columns 
flightnum <- select(flights, flight, year, month, day)

#Group by the year, month, day, (and flight). Count the number of times that flight number shows up that day.
flights_perday<- flightnum %>%
                 group_by(year, month, day, flight) %>%
                 count(flight)

print((flights_perday))
```

It looks like flight numbers were used more than once on any given day. If we only want to see the flight numbers that were reused on the same day then
```{r}
flight_dups <- flightnum %>% 
               count(year, month, day, flight) %>% 
               filter(n > 1)

print(flight_dups)

```


2. Add the location of the origin and destination (i.e. the lat and lon) to flights. Retain only columns for the airport code, airport name, longitude, and latitude. Name the columns in a user-friendly manner.

```{r}
airport_loc <- airports %>%
  select(faa, latitude = lat, longitude = lon, name = name)

flights_loc <- flights %>%
              select(origin, dest) %>%
              left_join(
                airport_loc,
                by = c("origin" = "faa")
                ) %>%
              left_join(
                airport_loc,
                by = c("dest" = "faa"),
                suffix = c("_origin", "_dest")
              )

flights_loc

```

3. Create appropriate numerical and/or graphical summaries to investigate how visibility condition makes it more likely to see a delay. Make sure to also investigate the flight delays with missing visibility values. (Hint: use left_join)


```{r}
#Known visibility values
#Departure delays
visibility_dep <- flights %>%
              left_join(weather) %>%
              mutate(visib_cat = cut_interval(visib, n = 10)) %>%
              group_by(visib_cat) %>%
              summarise(dep_delay = mean(dep_delay, na.rm = TRUE)) %>%
              ggplot(aes(x = visib_cat, y = dep_delay)) +
              geom_point() + ggtitle("Departure delay vs. Visibility") + 
              xlab("Visibility (miles)") + ylab("Departure delay (min)")
visibility_dep

#Arrival delays
visibility_arr <- flights %>%
                  left_join(weather) %>%
                  mutate(visib_cat = cut_interval(visib, n = 10)) %>%
                  group_by(visib_cat) %>%
                  summarise(arr_delay = mean(arr_delay, na.rm = TRUE)) %>%
                  ggplot(aes(x = visib_cat, y = arr_delay)) +
                  geom_point() + ggtitle("Arrival delay vs. Visibility") + 
                  xlab("Visibility (miles)") + ylab("Arrival delay (min)")
visibility_arr



#Missing visibility values 
#Departure delays
visib_missing_dep <- flights %>%
                     left_join(weather) %>%
                     filter(is.na(visib)) %>%
                     ggplot(aes(x = dep_delay)) + geom_histogram() 
visib_missing_dep


#Arrival delays
visib_missing_arr <- flights %>%
                     left_join(weather) %>%
                     filter(is.na(visib)) %>%
                     ggplot(aes(x = arr_delay)) + geom_histogram()
visib_missing_arr

```

If the visibility is less than 2 miles there is likely to be a delay (departure and/or arrival). There are approximately twice as many departure delays than arrival delays for the flights with missing visibility values.  

4. Which 3 destination airports have the most severe arrival delay? The severity of delay is defined as the proportion of arriving flights that have no less than 30 minutes arrival delay. Create a data frame containing their airport name, code, and the severity of delay.

```{r}
#Have to join with airport dataframe to include airport name
airport_name <- airports %>%
  select(faa, name = name)

severe_dest_delay <- flights %>%
                     select(dest, arr_delay) %>%
                     left_join(
                      airport_name,
                      by = c("dest" = "faa")
                     ) %>% 
                     mutate(delay = ifelse(arr_delay < 30, "normal", "severe")) %>%
                     count(dest, name, delay) %>% 
                     group_by(dest) %>%
                     mutate(prop = n / sum(n)) %>%
                     filter(delay == "severe") %>%
                     arrange(desc(prop), desc(n))

severe_dest_delay

```

The 3 destination airports with the most severe arrival delays are CAE (Columbia Metropolitan), TUL (Tulsa Intl), and OKC (Will Rogers World).  


5. What happened on June 13 2013? Look at the delay severity by airport on that day, and then use Google to cross-reference with the weather.


```{r}
jun_13 <- flights %>%
          filter(year == 2013, month == 6, day == 13) %>%
          group_by(dest) %>%
          summarise(delay = mean(arr_delay, na.rm = TRUE)) %>%
          inner_join(airports, by = c("dest" = "faa")) %>%
          arrange(desc(delay))

jun_13

```

On June 13, 2013 there was a series of derechos in the southeast. This explains why the airports located in the southeast had the most severe delays on that day. 